Reproducible Research: Peer Assessment 1
=========================================


## Loading and preprocessing the data
```{r input data and cleaning, tidy=FALSE}
library(lubridate)
activity <- read.csv("activity.csv",
                 stringsAsFactors = FALSE) # insure data is not converect to a factor
activity$date <- as.Date(activity$date)

str(activity) #show structure of data frame
```
## What is mean total number of steps taken per day?
```{r compute unadjusted steps, tidy=FALSE, fig.width=8,fig.height=8}
range <- as.numeric(max(activity$date)-min(activity$date))
daily <- data.frame(Date = as.Date(character()),
                   Steps = numeric(),
                   invalids = numeric(),
                   stringsAsFactors = FALSE)
base.date <- min(activity$date)
for(index in 0:range){
     filter.date <- base.date + index
     work <- subset(activity, date == filter.date)
     steps <- sum(work$steps,
                  na.rm = TRUE)
     invalids <- sum(is.na(work$steps))
     temp <- data.frame(Date = filter.date,
                       Steps = steps,
                       invalids = invalids,
                       stringsAsFactors = FALSE)
     daily <- rbind(daily,temp)
}
hist(daily$Steps,
     breaks = 12,
     main="Histogram of number of steps taken daily",
     xlab="Steps")
mean(daily$Steps)
median(daily$Steps)
```
As seen above the mean is `r mean(daily$Steps)` and the median is `r median(daily$Steps)`.
## What is the average daily activity pattern?
```{r daily pattern,tidy=FALSE,fig.width=10,fig.height=5}

range <- unique(activity$interval)
interval <- data.frame(interval = numeric(),
                       mean.steps = numeric(),
                       med.steps = numeric(),
                       std.steps = numeric(),
                       invalids = numeric())

for(index in seq_along(range)){
     filter <- range[index]
     work <- subset(activity, 
                    interval==filter,
                    select = steps)
     mean.steps <- mean(work$steps,
                        na.rm = TRUE)
     med.steps <- median(work$steps,
                         na.rm = TRUE)
     std.steps <- sd(work$steps,
                     na.rm = TRUE)
     invalids <- sum(is.na(work$steps))
     temp <- data.frame(interval = filter,
                        mean.steps = mean.steps,
                        med.steps = med.steps,
                        std.steps = std.steps,
                        invalids = invalids)
     interval <- rbind(interval,temp)
}

plot(x = interval$interval,
     y = interval$mean.steps,
     type = "l",
     main = "Mean number of steps per 5 minute time interval",
     xlab = "interval (5 minutes)",
     ylab = "steps")

interval$interval[which.max(interval$mean.steps)]
```
The 5-minute interval with the highest average number of steps is
`r interval$interval[which.max(interval$mean.steps)]`.
## Imputing missing values



## Are there differences in activity patterns between weekdays and weekends?
