Reproducible Research: Peer Assessment 1
=========================================


## Loading and preprocessing the data
The file activity.csv is read in from the working directory. 

The data from the file is placed into the activity data frame.

The _date_ varable inside the activity data frame is then converted to a date 
type variable.

```{r , tidy=FALSE}
# read the csv file in and assign to the activity data frome
activity <- read.csv("activity.csv",
                 stringsAsFactors = FALSE) # insure data is not converect to a factor
# convert the date field to a date data type
activity$date <- as.Date(activity$date)
str(activity)
summary(activity)
```
## What is mean total number of steps taken per day?
The activity data is now parsed and accumulated providing the total number of steps for each day,
this is stored in the daily data frame along with the number of missing observations
for each day.
```{r , tidy=FALSE}
#determin the number of days in the data set
range <- as.numeric(max(activity$date)-min(activity$date))

# set up a NULL data frame to hold the daily totals
daily <- data.frame(date = as.Date(character()),
                   steps = numeric(),
                   invalids = numeric(),
                   stringsAsFactors = FALSE)
# compute the total steps for a day and the number of NA values in that day
# add that to the daily data frame for each day
base.date <- min(activity$date)
for(index in 0:range){
     filter.date <- base.date + index
     work <- subset(activity, date == filter.date)
     steps <- sum(work$steps,
                  na.rm = TRUE)
     invalids <- sum(is.na(work$steps))
     temp <- data.frame(date = filter.date,
                       steps = steps,
                       invalids = invalids,
                       stringsAsFactors = FALSE)
     daily <- rbind(daily,temp)
}
```
Using the daily data agravated above a histogram of the number of steps taken on
a daily passes is now prepared. Also the mean and median number of steps is computed.
```{r,tidy=FALSE, fig.width=8, fig.height=8}
hist(daily$steps,
     breaks = 12,
     main="Histogram of number of steps taken daily",
     xlab="Steps")
# compute mean and median daily steps
mean(daily$steps)
median(daily$steps)
```
As seen above the mean is **`r mean(daily$steps)`** steps and the median is 
**`r median(daily$steps)`** steps.
## What is the average daily activity pattern?
The activity data is now agravated by the spcified interval, or 5 minute period
in the day. FOr each interval the average (mean) number of steps is computed
along with the median, standard deviation and the number of missing values for each
interval are computed and stored inthe interval data frame.
```{r,tidy=FALSE,fig.width=10,fig.height=5}
# get a list of valid intervals
range <- unique(activity$interval)

#set up NULL interval data frame
interval <- data.frame(interval = numeric(),
                       mean.steps = numeric(),
                       med.steps = numeric(),
                       std.steps = numeric(),
                       invalids = numeric())

# compute average steps per interval and stor in interval data frame
# additionally the median and standard deviation is computed for each interval
# as well as the number of NA for each interval.
for(index in seq_along(range)){
     filter <- range[index]
     work <- subset(activity, 
                    interval==filter,
                    select = steps)
     mean.steps <- mean(work$steps,
                        na.rm = TRUE)
     med.steps <- median(work$steps,
                         na.rm = TRUE)
     std.steps <- sd(work$steps,
                     na.rm = TRUE)
     invalids <- sum(is.na(work$steps))
     temp <- data.frame(interval = filter,
                        mean.steps = mean.steps,
                        med.steps = med.steps,
                        std.steps = std.steps,
                        invalids = invalids)
     interval <- rbind(interval,temp)
}
```
Using the data agervated above a plot is now prepared to show the number of steps
ddistrubation accross the five minute intervals for a day. As can be expected
ther are periods of very little activity and smaller periods of very high levels
of activity.
```{r,tidy=FALSE,fig.width=10,fig.height=5}
plot(x = interval$interval,
     y = interval$mean.steps,
     type = "l",
     main = "Mean number of steps per 5 minute time interval",
     xlab = "interval (5 minutes)",
     ylab = "steps")
```
The 5-minute interval with the highest activity is identify.
```{r}
interval$interval[which.max(interval$mean.steps)]
```
The 5-minute interval with the highest average number of steps is
**`r interval$interval[which.max(interval$mean.steps)]`**.
## Imputing missing values
The data is now divided into two data frame with the observations that are missing
a step count being placed into one and ovbservations with a step count being placed into 
another data frame. A review of the data showed that only the steps varable had missing data.

```{r, tidy=FALSE}

good <- subset(activity,
               subset = !is.na(steps))
bad <- subset(activity,
              subset = is.na(steps))
```
The number of observation that were missing the number of steps are counted.
```{r, tidy=FALSE}
#number of NA observatins
nrow(bad)
```
There were `r nrow(bad)` observations without the number of steps.

The missing observations will be replace with the mean number of steps for that 
interval. The mean number of steps will be obtained from the calculations above
that was stored inthe interval data frame. The two data frames are then recombined 
into a single data frame.
```{r, tidy=FALSE}
# the number of observations needing correction is assigned
range = nrow(bad)
# replace the NA values with the mean value for that interval.
for (index in 1:range){
     filter <- bad$interval[index]
     bad$steps[index] <- subset(interval, interval == filter)$mean.steps
}

#combine the two data frames
activity_imput <- rbind(good,bad)
```
The number of steps taken in a day is now recomputed and stored in a new data
frame. This new data frame is then used to replot the histograme of daily steps.
Finally a new mean and median are computed for the imputed data and compaired to
the data with the missing data was ignored.

```{r, tidy=FALSE, fig.width=10, fig.height=5}
# set number of days in the data set
range <- as.numeric(max(activity_imput$date)-min(activity_imput$date))

# set up NULL data frame for results
daily_imput <- data.frame(date = as.Date(character()),
                          steps = numeric(),
                          invalids = numeric(),
                          stringsAsFactors = FALSE)

# collect and total number of steps per day
base.date <- min(activity_imput$date)

for(index in 0:range){
     filter.date <- base.date + index
     work <- subset(activity_imput, date == filter.date)
     steps <- sum(work$steps,
                  na.rm = TRUE)
     invalids <- sum(is.na(work$steps))
     temp <- data.frame(date = filter.date,
                        steps = steps,
                        invalids = invalids,
                        stringsAsFactors = FALSE)
     daily_imput <- rbind(daily_imput,temp)
}

# plot histogram
hist(daily_imput$steps,
     breaks = 12,
     main="Histogram of number of steps taken daily
     with missing data points changed to the mean",
     xlab="Steps")

# compute new mean and median
mean(daily_imput$steps)
median(daily_imput$steps)

```
The mean total steps per day using the adjusted data is **`r mean(daily_imput$steps)`**
a difference of **`r mean(daily_imput$steps)-mean(daily$steps)`** steps from the 
unadjusted mean of **`r mean(daily$steps)`** steps.

The median total steps per day using the adjusted data is **`r median(daily_imput$steps)`**
a difference of **`r median(daily_imput$steps)-median(daily$steps)`** steps from the
unadjusted median of **`r median(daily$steps)`** steps.
## Are there differences in activity patterns between weekdays and weekends?
THe imputted data (data were the missing observations has been replaced), is now used
to see if there are varing activaty patterns between weekdays and weekends.
The data is frist classiffied as to weather a date is a weekday or weekend
```{r,tidy=FALSE,fig.width=8}
range = nrow(activity_imput)
work <- data.frame(weekend = factor(levels=c("weekday","weekend")),
                   stringsAsFactors = FALSE)

for(index in 1:range){
     hold <- weekdays(activity_imput$date[index])
     if (hold == "Saturday" | hold =="Sunday"){
          weekend = "weekend"
     } else {
          weekend = "weekday"
     }
     temp <- data.frame(weekend = as.factor(weekend),
                        stringsAsFactors = FALSE)
     work <- rbind(work,temp)     
}
activity_imput  <- cbind(activity_imput,work)
```
Once the day is classified the data frame is devided into two (weekday and weekend),
and new average steps per interval are computed for each data frame. 
```{r,tidy=FALSE,fig.width=8}
# plot weekdays
range <- unique(activity_imput$interval)
temp_weekday <- subset(activity_imput,
                       weekend == "weekday")
interval_weekday <- data.frame(interval = numeric(),
                               mean.steps = numeric(),
                               weekend = factor(levels=c("weekday","weekend")),
                               stringsAsFactors = FALSE)
temp_weekend <- subset(activity_imput,
                       weekend == "weekend")
interval_weekend <- data.frame(interval = numeric(),
                               mean.steps = numeric(),
                               weekend = factor(levels=c("weekday","weekend")),
                               stringsAsFactors = FALSE)


for(index in seq_along(range)){
     filter <- range[index]
     work <- subset(temp_weekday, 
                    interval==filter)
     
     mean.steps <- mean(work$steps,
                        na.rm = TRUE)

     temp <- data.frame(interval = filter,
                        mean.steps = mean.steps,
                        weekend = as.factor(work$weekend))
     interval_weekday <- rbind(interval_weekday, temp)
}

for(index in seq_along(range)){
     filter <- range[index]
     work <- subset(temp_weekend, 
                    interval==filter)
     
     mean.steps <- mean(work$steps,
                        na.rm = TRUE)

     temp <- data.frame(interval = filter,
                        mean.steps = mean.steps,
                        weekend = as.factor(work$weekend))
     interval_weekend <- rbind(interval_weekend, temp)
}
```
The average activity per interval is now plotted for each data frame, showing
differing activity patterns for weekday and weekend activity throughout
the day

```{r,tidy=FALSE,fig.width=8}
par(mfrow=(c(2,1)))

plot(x = interval_weekday$interval,
     y = interval_weekday$mean.steps,
     type = "l",
     main = "Mean number of steps per 5 minute time interval
     adjusted values
     Weekdays",
     xlab = "interval (5 minutes)",
     ylab = "steps")
              
plot(x = interval_weekend$interval,
     y = interval_weekend$mean.steps,
     type = "l",
     main = "Mean number of steps per 5 minute time interval
     adjusted values
     Weekends",
     xlab = "interval (5 minutes)",
     ylab = "steps")
par(mfrow=(c(1,1)))     
```

